---
title: "Data Dictionary"
author: "Zachary M. Smith"
format: 
  html:
    page-layout: full
---

The data dictionary below describes data from the WQMA Oracle database and the WQMA Analytical Data Store (WQMA parquet files). To get this table in R, use:

```{r}
data("columns_dictionary", package = "nexus")
```

::: callout-important
The data dictionary empowers data users to quickly become familiar with a data set. Although the table is simple, it is critical to operations that this table remain up-to-date, accurate, and easy to understand.

If you find any errors or do not find the documentation clear, please open a [GitHub Issue](https://github.com/BWAM/bwam_analytics/issues) and we can collectively work towards a solution.
:::

The dictionary provides:

1.  [**table_name:**]{.underline} The name of the table in the WQMA Oracle Database where the **column_name** is stored.

2.  [**column_name:**]{.underline} The name of the column in the data store of interest. The naming convention is the same between the WQMA Oracle Database and WQMA Analytical Data Store.

3.  [**definition:**]{.underline} A description of what the column represents.

4.  [**examples:**]{.underline} Provides examples of the data found in a column. If more than 20 unique values are represented in the column, then 20 example character strings will be provided. If 20 or less unique values are represented in the column, all 20 values are provided.

```{r, echo=FALSE}

library(DT)
data("columns_dictionary",
     package = "nexus")

DT::datatable(
  columns_dictionary |> 
    dplyr::select(table_name,
                  column_name,
                  definition,
                  examples
                  ),
  filter = 'top',
  extensions = c('Buttons',
                 'FixedHeader'),
  
  options = list(pageLength = 5,
                 # autoWidth = TRUE,
                 fixedHeader = TRUE,
                 dom = 'Bfrtip',
                 buttons = c('copy',
                             'csv',
                             'excel',
                             'pdf',
                             'print'),
                 columnDefs = list(
    list(width = '25px', targets = c(1, 2)),
    list(width = '100px', targets = c(3, 4)))
    )
) |> 
  DT::formatStyle(columns = 1:ncol(columns_dictionary),
                  fontSize = '75%')

```
