---
title: "WQX API"
author: "Amanda Herbst"
format: html
---

The Water Quality eXchange (WQX) API allows us to directly connect to WQXWeb and facilitate data uploads without having to manually do so within WQXWeb.

## Initial Requirements

In order to do so, you must have a CDX account and generate a private encryption key because the process checks for your credentials:

#### 1. Create CDX account

-   Follow [these instructions](https://www.epa.gov/waterdata/wqx-web-account-registration) to create an account.
    -   WQX Organization ID: `21NYDECA_WQX`
    -   WQX Organization Name: `New York State Dec Division of Water`

#### 2. Generate Private Encryption Key

-   Once you are logged into WQXWeb, go to "Setup" → "My User Account Details" and click the "Create New Key" button at the bottom of the page.
-   Copy this key to your clipboard and **click save**.
-   visit [here](https://www.epa.gov/waterdata/wqx-web-application-programming-interface) for more info

## Setup

Ensure the following packages are installed

```{r, eval = FALSE}
library(httr2) # for working with the WQXWeb API
library(keyring) # for securely storing the private encryption key
library(digest) # for encrypting your credentials
library(jsonlite) # for encryption and handling of data 
library(glue) # for concatonating strings
```

First, store your private encryption key by running the following. A window will pop up, prompting you to paste your key. Once you paste and hit okay, your key will be stored under the name "WQXWeb".

```{r, eval = FALSE}
keyring::key_set(service = "WQXWeb")
```

Define some universal variables

-   `user_id`: your username for logging into CDX, typically FirstnameLastname
-   `organization_id`: this is DEC's ID in WQX
-   `base_uri`: the base Uniform Resource Identifier, essentially the first part of the url to the api that will be used every time

```{r, eval = FALSE}
user_id <- "YourUserID"

organization_id <- "21NYDECA_WQX"

base_uri <- "https://cdx.epa.gov/WQXWeb/api/"
```

## WQX Request Function

This function will be used to perform all service calls to the WQXWeb API, whether it is uploading data or pulling data.

Every call requires a header that contains:

-   `X-UserID`: the user ID as defined above
-   `X-Stamp`: the timestamp when the request is made in UTC Time in the format `mm/dd/yyyy hh:mi:ss AM`
-   `X-Signature`: a concatenated string made up of the User ID, Timestamp, URI, and Request Method that is encrypted using the HMAC-SHA256 encryption algorithm
-   `Content-Type`: the format in which you would like the response, here we chose JSON

```{r, eval = FALSE}
wqx_request <- function(request_method, user_id, uri, data_file_path = NULL){
  
  # type of content returned
  contenttype <- 'application/json'
  
  private_key <- jsonlite::base64_dec(keyring::key_get(service = "WQXWeb"))
  
  # time must be UTC and in format mm/dd/yyyy hh:mi:ss AM
  # %I is standard time 01-12 hours, while %H is military time 01-24 hours
  timestamp <- format(as.POSIXct(Sys.time(), tz = "UTC"), "%m/%d/%Y %I:%M:%S %p") 
  
  # make signature
  signature <- paste0(user_id, timestamp, uri, request_method) 
  
  # encrypt signature
  encrypted_sig <- digest::hmac(key = private_key, object = signature, algo = "sha256", raw = T)|> 
    jsonlite::base64_enc() 
  
  
  if (request_method == "POST") {
    r <- request(uri) |>
      req_headers("X-UserID" = user_id,
                  "X-Stamp" = timestamp,
                  "X-Signature" = encrypted_sig,
                  "Content-Type" = contenttype) |>
      req_body_file(data_file_path) 
    
    r |>
      req_perform()
  }
  
  if (request_method == "GET") {
    r <- request(uri) |>
      req_headers("X-UserID" = user_id,
                  "X-Stamp" = timestamp,
                  "X-Signature" = encrypted_sig,
                  "Content-Type" = contenttype)
    r |>
      req_perform()
  }
}
```

## Service Calls

### Upload

*Upload a file to the web server (to be imported).*

Define the path to the file to be uploaded. Then, the filename will be extracted and used to complete the URI for this service call.

```{r, eval = FALSE}
file_path <- "file/path/to/data"

filename <- basename(file_path)

uri_upload <- glue::glue("{base_uri}Upload/{filename}")
```

Make the service call:

```{r, eval = FALSE}
resp <- wqx_request(request_method = "POST",
                    user_id = user_id,
                    uri = uri_upload,
                    data_file_path = file_path)
```

### StartImport

*Start importing the file that was just uploaded in the previous step*

### StartXmlExport

*Create the XML submission file that goes to CDX*

### SubmitDatasetToCdx

*Submit the data set to CDX*

### SubmitFileToCdx

*Submit a preciously uploaded WQX XML file to CDX*

### GetStatus

*Get the status for a specific dataset For large imports (longer than 20 minutes), calling this service periodically will guarantee that the server will not shutdown before the import completes.*

### GetDocumentList

*Get the list of available documents for a dataset*

### Projects

*Get all the projects for our organization that exist in WQXWeb.*

```{r, eval = FALSE}
uri_projects <- glue::glue("{base_uri}Projects?OrganizationIdentifiersCsv={organization_id}")

resp <- wqx_request(request_method = "GET",
                    user_id = user_id,
                    uri = uri_projects)

# coerce response into dataframe
projects_df <- resp |> resp_body_string() |> fromJSON(flatten = TRUE)
```

### MonitoringLocations

*Get all the monitoring locations for our organization that exist in WQXWeb*
